---
title: "Stock Market Returns Distributions"
subtitle : 
author   : Giovanni Fossati
job      : null
output   : 
  html_document:
    self_contained: false
    css: css/gf_new_nolines.css
    theme: cerulean
    highlight: tango
    keep_md: no
    mathjax: default
    toc: no
    includes:
      md_extensions: +fenced_code_attributes
---

```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
require("knitr")
options(width = 100, 
        scipen = 5)
opts_chunk$set(message = FALSE, 
               error = FALSE, 
               warning = FALSE, 
               collapse = TRUE, 
               tidy = FALSE,
               cache = FALSE, 
               cache.path = '.cache/', 
               comment = '#',
               fig.align = 'center', 
               dpi = 100, 
               dev = "png",
               fig.path = 'figures/')
```

## PRELIMINARIES

Libraries needed for data processing and plotting:

```{r load_packages, cache = FALSE, echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE}
# library("lubridate")
# library("plyr")
library("dplyr")
library("tidyr")
library("ggplot2")
library("gridExtra")
```

<script src="./lib/toggle_GF.js"></script>

## Introduction

Two interesting, blog posts:

* [The Generalized Lambda Distribution and GLDEX Package: Fitting Financial Return Data](http://www.r-bloggers.com/the-generalized-lambda-distribution-and-gldex-package-fitting-financial-return-data/)
* [The Generalized Lambda Distribution and GLDEX Package for Fitting Financial Return Data – Part 2](http://www.r-bloggers.com/the-generalized-lambda-distribution-and-gldex-package-for-fitting-financial-return-data-part-2/)

The GLDEX package is available on [CRAN](http://cran.r-project.org/web/packages/GLDEX/index.html)


<a name="THE_DATA"></a>

## Libraries and Other Setup

```{r load-libraries, echo = TRUE, cache = FALSE}
library("quantmod")
library("GLDEX")
library("MASS")
```

```{r load-more_libraries_for_plotting, echo = TRUE, cache = FALSE}
library("scales")
source("./scripts/my_functions.R")
```

```{r set_dates}
start.date <- "1986-01-01"
end.date <- "2014-12-31"

xxlin <- seq(-25.0, 25.0, by = 0.1)
xxlog <- seq(-2.0, 2.0, by = 0.001)
```

<hr class="thin_separator">

# Three Broad Market Indices : Daily Returns

We will look at data of three broad market indices to get an idea of the movements of the market as a whole
and obtain distributions/statistics to use in the _Rent vs. Buy_ application as the basis of the simulated 
market returns.

## Returns

We consider two (obviously related) quantities expressing the return of an asset described by an ordered set of
values ${x_i}$:

* _log return_: $$ {rl}_i = \log\left(\frac{x_i}{x_{i-1}}\right) $$

* _relative return_: $$ {rr}_i = e^{rl_i} - 1 = \frac{x_i - x_{i-1}}{x_{i-1}} $$

## Wilshire 5000

Fetch the data with `quantmod`:

```{r wilshire5000-load, cache = TRUE}
# getSymbols("VTSMX", from = "1994-01-01")
getSymbols("VTSMX", from = start.date, to = end.date)
VTSMX.vec <- as.vector(VTSMX[, 4])
```

Prepare returns:

```{r wilshire5000-prepare_data, cache = TRUE}
WW <- prepare_data(data = VTSMX.vec, xlin = xxlin, xlog = xxlog)

WW$dates <- index(VTSMX)
WW$year <- substr(index(VTSMX), 1, 4)
```

The function `prepare_data()` returns a list with (for each _rr_ and _rl_):
the data themselves, 
parameters of the best fit with the generalized Gamma distribution,
distributions for this latter and two Gaussians fit to the data in two narrow (less than $\sigma$ wide) 
ranges around the mean.   
We add to this list dates from the timeseries _index_ and year:

This is the structure of the resulting list:

<button class="toggle_code">show output</button>
```{r wilshire5000-show_prepared_data_list, echo = TRUE}
str(WW)
```
 
```{r wilshire5000-plot_distributions_plain, echo = FALSE, cache = TRUE, eval = FALSE}
# NOT EVALUATED
hist(WW$rr_data, breaks = seq(-55.0, 55.0, by = 0.10), xlim = c(-8.0, 8.0), freq = FALSE,
     xlab = "daily returns (%)",
     main = "Wilshire 5000")
lines(WW$rr_distr$x, WW$rr_distr$y_norm1, col = "green2", lwd = 2)
lines(WW$rr_distr$x, WW$rr_distr$y_norm2, col = "blue2",  lwd = 2)
lines(WW$rr_distr$x, WW$rr_distr$y_gl,    col = "red2",   lwd = 2)
```

#### Distributions of _relative returns_ and of _log returns_:

The histograms are the return data, with overlayed 

* in red the _generalized Gamma distribution_, and 
* in blue the two Gaussians fit to the "core" of the data.

It is quite obvious how much better the _generalized Gamma distribution_ describes the data, 
both their "code" and the long fat tails.

<button class="toggle_plot_code">show plot code</button>
```{r wilshire5000-plot_distributions_ggplot_3, echo = FALSE, cache = FALSE, fig.width = 10.0, fig.height = 5.0}
p1 <- plot_distr_rel_returns(data = WW$rr_data, fits = WW$rr_distr)
p2 <- plot_distr_log_returns(data = WW$rl_data, fits = WW$rl_distr)
grid.arrange(p1, p2, nrow=1)
```


<hr class="thin_separator">

### S&P 500 

Some treatment as the _Wilshire 5000_ data.

Fetch the data with `quantmod`:

```{r sp500-load, cache = TRUE}
getSymbols("^GSPC", from = start.date, to = end.date)
GSPC.vec <- as.vector(GSPC[, 4])
```

and prepare with the `prepare_data()` function:

```{r sp500-prepare_data, cache = TRUE}
SP <- prepare_data(data = GSPC.vec, xlin = xxlin, xlog = xxlog)

SP$dates <- index(GSPC)
SP$year <- substr(index(GSPC), 1, 4)
```

This is the structure of the resulting list:

<button class="toggle_code">show output</button>
```{r sp500-show_prepared_data_list, echo = TRUE}
str(SP)
```
 

#### Distributions of _relative returns_ and of _log returns_:

<button class="toggle_plot_code">show plot code</button>
```{r sp500-plot_distributions_ggplot_3, echo = TRUE, cache = FALSE, fig.width = 10.0, fig.height = 5.0}
p1 <- plot_distr_rel_returns(data = SP$rr_data, fits = SP$rr_distr)
p2 <- plot_distr_log_returns(data = SP$rl_data, fits = SP$rl_distr)
grid.arrange(p1, p2, nrow=1)
```


<hr class="thin_separator">

### NASDAQ 100

Fetch the data with `quantmod`:

```{r nasdaq100-load, cache = TRUE}
getSymbols("^NDX", from = start.date, to = end.date)
NDX.vec <- as.vector(NDX[, 4])
```

```{r nasdaq100-prepare_data, cache = TRUE}
NN <- prepare_data(data = NDX.vec, xlin = xxlin, xlog = xxlog)

NN$dates <- index(NDX)
NN$year <- substr(index(NDX), 1, 4)
```

This is the structure of the resulting list:

<button class="toggle_code">show output</button>
```{r nasdaq100-show_prepared_data_list, echo = TRUE}
str(NN)
```
 
#### Distributions of _relative returns_ and of _log returns_:

<button class="toggle_plot_code">show plot code</button>
```{r nasdaq100-plot_distributions_ggplot_3, echo = TRUE, cache = FALSE, fig.width = 10.0, fig.height = 5.0}
p1 <- plot_distr_rel_returns(data = NN$rr_data, fits = NN$rr_distr)
p2 <- plot_distr_log_returns(data = NN$rl_data, fits = NN$rl_distr)
grid.arrange(p1, p2, nrow=1)
```


<hr class="thin_separator">

# Three Broad Market Indices : Annual Returns


## Annual from Daily Returns

Simulating annual returns by compounding daily returns drawn from the 
best fit _generalized gamma distributions_.

<button class="toggle_code">show code</button>
```{r annual_returns_simulated_1, eval = TRUE, cache = TRUE, echo = TRUE}
# typical number of market days in a year is ~252
ndays <- 252

# repeats
  N <- 1e5
# N <- 1e6

set.seed(1212)
test_rr.W5000 <- rep(0, N)
for(j in 1:N) {
    dummy <- rgl(ndays, WW$rr_fit_gl)
    test_rr.W5000[j] <- (cumprod(1 + dummy/100)[ndays] - 1.0)*100
}

test_rr.SP500 <- rep(0, N)
for(j in 1:N) {
    dummy <- rgl(ndays, SP$rr_fit_gl)
    test_rr.SP500[j] <- (cumprod(1 + dummy/100)[ndays] - 1.0)*100
}

test_rr.N100 <- rep(0, N)
for(j in 1:N) {
    dummy <- rgl(ndays, NN$rr_fit_gl)
    test_rr.N100[j] <- (cumprod(1 + dummy/100)[ndays] - 1.0)*100
}
```

<button class="toggle_code">show code</button>
```{r annual_returns_simulated_from_gaussian, eval = TRUE, cache = TRUE, echo = TRUE}
# typical number of market days in a year is ~252
ndays <- 252

# repeats
  N <- 1e5
# N <- 1e6

set.seed(1414)
test_rr_n1.W5000 <- rep(0, N)

print(ndays)
print(WW$rr_fit_n1)

for(j in 1:N) {
    dummy <- rnorm(ndays, WW$rr_fit_n1)
    test_rr_n1.W5000[j] <- (cumprod(1 + dummy/100)[ndays] - 1.0)*100
}

test_rr_n1.SP500 <- rep(0, N)
for(j in 1:N) {
    dummy <- rnorm(ndays, SP$rr_fit_n1)
    test_rr_n1.SP500[j] <- (cumprod(1 + dummy/100)[ndays] - 1.0)*100
}

test_rr_n1.N100 <- rep(0, N)
for(j in 1:N) {
    dummy <- rnorm(ndays, NN$rr_fit_n1)
    test_rr_n1.N100[j] <- (cumprod(1 + dummy/100)[ndays] - 1.0)*100
}
```

<button class="toggle_plot_code">show plot code</button>
```{r plot-annual_returns_distr_1_rr, echo = TRUE, eval = TRUE, fig.width = 8.0, fig.height = 5.5}
xx2 <- seq(-200, 200, by = 1.0)
yy.W5000.n <- dnorm(xx2, mean = mean(test_rr.W5000), sd = sd(test_rr.W5000))
yy.SP500.n <- dnorm(xx2, mean = mean(test_rr.SP500), sd = sd(test_rr.SP500))
yy.N100.n  <- dnorm(xx2, mean = mean(test_rr.N100), sd = sd(test_rr.N100))

plot(density(test_rr.N100), type = "n",
     xlim = c(-100.0, 200.0), 
     ylim = c(0.0, 0.037), 
     xlab = "annual return [%]", 
     main = "Wilshire5000, S&P500, NASDAQ100")
grid()
lines(xx2, yy.N100.n,         col = "red2",   lwd = 3, lty = 3)
lines(xx2, yy.SP500.n,        col = "blue2",  lwd = 3, lty = 3)
lines(xx2, yy.W5000.n,        col = "green3", lwd = 3, lty = 3)
lines(density(test_rr.N100),  col = "red2",   lwd = 3)
lines(density(test_rr.SP500), col = "blue2",  lwd = 3)
lines(density(test_rr.W5000), col = "green3", lwd = 3)
legend("topleft", bty = "y", bg = "grey95", x.intersp = 0.7, y.intersp = 0.8, 
       legend = c("Nasdaq100", "S&P 500", "Wilshire5000"), 
       lty = 1, lwd = 3,
       col = c("red2", "blue2", "green3"),
       cex = 1.0)
legend("topright", bty = "y", bg = "grey95", x.intersp = 0.7, y.intersp = 0.8, 
       legend = c("from simulations", "Gaussian"),
       lty = c(1, 3), lwd = 3, seg.len = 4,
       col = c("grey10", "grey10"),
       cex = 1.0)
```

<button class="toggle_plot_code">show plot code</button>
```{r plot-annual_returns_distr_2_from_gaussian, echo = FALSE, eval = FALSE, fig.width = 8.0, fig.height = 5.5}
# NOT EVALUATED ('test_rl' DATA NOT PREPARED)
plot(density(test_rl.N100), type = "n",
     xlim = c(-100.0, 200.0), 
     ylim = c(0.0, 0.037), 
     xlab = "annual return [%]", 
     main = "Wilshire5000, S&P500, NASDAQ100")
grid()
lines(density(test_rr_n1.N100),  col = "red2",   lwd = 3, lty = 3)
lines(density(test_rr_n1.SP500), col = "blue2",  lwd = 3, lty = 3)
lines(density(test_rr_n1.W5000), col = "green3", lwd = 3, lty = 3)
lines(density(test_rl.N100),  col = "red2",   lwd = 3)
lines(density(test_rl.SP500), col = "blue2",  lwd = 3)
lines(density(test_rl.W5000), col = "green3", lwd = 3)
legend("topleft", bty = "y", bg = "grey95", x.intersp = 0.7, y.intersp = 0.8, 
       legend = c("Nasdaq100", "S&P 500", "Wilshire5000"), 
       lty = 1, lwd = 3,
       col = c("red2", "blue2", "green3"),
       cex = 1.0)
legend("topright", bty = "y", bg = "grey95", x.intersp = 0.7, y.intersp = 0.8, 
       legend = c("from simulations", "Gaussian"),
       lty = c(1, 3), lwd = 3, seg.len = 4,
       col = c("grey10", "grey10"),
       cex = 1.0)
```

```{r plot-annual_returns_distr_3_rr_vs_rl, echo = FALSE, eval = FALSE, fig.width = 8.0, fig.height = 5.5}
# NOT EVALUATED
plot(density(test_rl.N100), type = "n",
     xlim = c(-100.0, 200.0), 
     ylim = c(0.0, 0.037), 
     xlab = "annual return [%]", 
     main = "Wilshire5000, S&P500, NASDAQ100")
grid()
lines(density(test_rl.N100),  col = "red2",   lwd = 3)
lines(density(test_rl.SP500), col = "blue2",  lwd = 3)
lines(density(test_rl.W5000), col = "green3", lwd = 3)
lines(density(test_rr.N100),  col = "red2",   lwd = 3, lty = 3)
lines(density(test_rr.SP500), col = "blue2",  lwd = 3, lty = 3)
lines(density(test_rr.W5000), col = "green3", lwd = 3, lty = 3)
```

## Annual straight from data

The simulated daily-compounded annual returns can be compared with a crude estimate
of annual returns obtained straight from the data by looking at 252-day-lagged values.

For this analysis we restricted the data to the 1994-2014 interval, in order to have
the same coverage for the three indices.

<button class="toggle_code">show code</button>
```{r aligned_annual_returns, echo = TRUE}
WWb <- VTSMX.vec[WW$year >= 1994]
SPb <- GSPC.vec[SP$year >= 1994]
NNb <- NDX.vec[NN$year >= 1994]
WWd <- index(VTSMX)[WW$year >= 1994]
SPd <- index(GSPC)[SP$year >= 1994]
NNd <- index(NDX)[NN$year >= 1994]

W5000_rla <- diff(log(WWb), lag = 252)
W5000_rra <- 100.0*(exp(W5000_rla)-1)

SP500_rla <- diff(log(SPb), lag = 252)
SP500_rra <- 100.0*(exp(SP500_rla)-1)

N100_rla <- diff(log(NNb), lag = 252)
N100_rra <- 100.0*(exp(N100_rla)-1)

rr.df <- data.frame(date = WWd[-(1:252)], WW = W5000_rra, SP = SP500_rra, NN = N100_rra)
rr.df$time <- as.POSIXct(rr.df$date)

rr.df$WWidx <- WWb[-(1:252)]
rr.df$SPidx <- SPb[-(1:252)]
rr.df$NNidx <- NNb[-(1:252)]

rr.df$WWidx_n <- WWb[-(1:252)]/rr.df$WWidx[rr.df$date == "1996-01-02"]
rr.df$SPidx_n <- SPb[-(1:252)]/rr.df$SPidx[rr.df$date == "1996-01-02"]
rr.df$NNidx_n <- NNb[-(1:252)]/rr.df$NNidx[rr.df$date == "1996-01-02"]
```

This crude estimates of annual returns are plotted here with dotted distributions, which exhibit a 
significant degree of irregularity. 
In particular all indices, being all stock-based, show similar double humped structure with a main hump at
moderate positive returns and a smaller one at around 10-40% negative returns.

These two humps are somewhat disconnected, reflecting the presence of almost two different market behaviors
in the last 20 years, with two period of major losses almost isolated in the midst of a long term
steady-ish rising trend (see plots further down).


<button class="toggle_plot_code">show plot code</button>
```{r plot-annual_returns_distr_2, echo = TRUE, eval = TRUE, fig.width = 8.0, fig.height = 5.5}
plot(density(W5000_rra), type = "n", 
     xlim = c(-100.0, 200.0), 
     ylim = c(0.0, 0.037), 
     xlab = "annual return [%]", 
     main = "Wilshire5000, S&P500, NASDAQ100")
grid()
lines(density(N100_rra),      col = "red2",   lwd = 3, lty = 3)
lines(density(SP500_rra),     col = "blue2",  lwd = 3, lty = 3)
lines(density(W5000_rra),     col = "green3", lwd = 3, lty = 3)
lines(density(test_rr.N100),  col = "red2",   lwd = 3)
lines(density(test_rr.SP500), col = "blue2",  lwd = 3)
lines(density(test_rr.W5000), col = "green3", lwd = 3)
legend("topleft", bty = "y", bg = "grey95", x.intersp = 0.7, y.intersp = 0.8, 
       legend = c("Nasdaq100", "S&P 500", "Wilshire5000"), 
       lty = 1, lwd = 3,
       col = c("red2", "blue2", "green3"),
       cex = 1.0)
legend("topright", bty = "y", bg = "grey95", x.intersp = 0.7, y.intersp = 0.8, 
       legend = c("from simulations", "straight from data"),
       lty = c(1, 3), lwd = 3, seg.len = 4,
       col = c("grey10", "grey10"),
       cex = 1.0)
```

#### Time-series plots of returns

```{r plot-annual_returns_timeseries_ggplot, echo = FALSE, eval = FALSE, fig.width = 10.0, fig.height = 5.5}
# NOT EVALUATED
ggplot(data = rr.df, aes(x = time)) + theme_bw() + 
       theme(axis.title = element_text(size = 16),
             axis.text= element_text(size = 14),
             axis.line = element_line(size = 1)) +
       coord_cartesian(ylim = c(-75.0, 125.0)) +
       xlab("date") + 
       ylab("annualized returns") + 
       scale_x_date(minor_breaks = "1 year") +
       geom_hline(h=0, lty = 2, col = "orange") +
       geom_line(aes(x = date, y = NN), col = "red2") + 
       geom_line(aes(x = date, y = SP), col = "blue2") + 
       geom_line(aes(x = date, y = WW), col = "green4")
```

<button class="toggle_plot_code">show plot code</button>
```{r plot-annual_returns_timeseries_ggplot_ALT, echo = TRUE, fig.width = 10.0, fig.height = 5.5}
tmp <- dplyr::select(rr.df, date, time, WW, SP, NN) %>% 
              gather(. , name, value, NN, SP, WW)

ggplot(data = tmp, aes(x = date, y = value)) + theme_bw() + 
       theme(axis.title = element_text(size = 16),
             axis.text= element_text(size = 14),
             axis.line = element_line(size = 1),
             legend.position = c(0.9, 0.85)) +
       coord_cartesian(ylim = c(-75.0, 135.0)) +
       xlab("date") + 
       ylab("annualized returns") + 
       geom_hline(h=0, lty = 2, col = "orange") +
       geom_line(aes(color = name)) + 
       scale_colour_manual(values = c("red2", "blue2", "green4"), name = "Index", 
                           labels = c("NASDAQ100", "S&P500", "Wilshire5000"))
```

```{r plot-indices_timeseries_ggplot, echo = FALSE, eval = FALSE, fig.width = 10.0, fig.height = 5.5}
# NOT EVALUATED
ggplot(data = rr.df, aes(x = time)) + theme_bw() + 
       theme(axis.title = element_text(size = 16),
             axis.text= element_text(size = 14),
             axis.line = element_line(size = 1)) +
       xlab("date") + 
       ylab("indices") + 
       scale_x_date(minor_breaks = "1 year") +
       scale_y_log10() +  
       geom_line(aes(x = date, y = NNidx), col = "red2") + 
       geom_line(aes(x = date, y = SPidx), col = "blue2") + 
       geom_line(aes(x = date, y = WWidx), col = "green4")
```

#### Time-series plots of indices (normalized to 1996-01-02)

```{r plot-indices_timeseries_ggplot_norm, echo = FALSE, eval = FALSE, fig.width = 10.0, fig.height = 5.5}
# NOT EVALUATED

ggplot(data = rr.df, aes(x = time)) + theme_bw() + 
       theme(axis.title = element_text(size = 16),
             axis.text= element_text(size = 14),
             axis.line = element_line(size = 1)) +
       xlab("date") + 
       ylab("indices (scaled to 1996-01-02)") + 
       scale_x_date(minor_breaks = "1 year") +
       scale_y_log10(breaks = c(1, 2, 4, 8)) +  
       geom_hline(h = 0, lty = 2, col = "orange") +
       geom_vline(xintercept = as.numeric(as.Date("1996-01-02")), lty = 2, col = "orange") +
       geom_line(aes(x = date, y = NNidx_n), col = "red2", show_guide = TRUE) + 
       geom_line(aes(x = date, y = SPidx_n), col = "blue2", show_guide = TRUE) + 
       geom_line(aes(x = date, y = WWidx_n), col = "green4")
```

<button class="toggle_plot_code">show plot code</button>
```{r plot-indices_timeseries_ggplot_norm_ALT, echo = TRUE, eval = TRUE, fig.width = 10.0, fig.height = 5.5}
tmp <- dplyr::select(rr.df, date, time, WWidx_n, SPidx_n, NNidx_n) %>% 
              gather(. , name, value, NNidx_n, SPidx_n, WWidx_n)

ggplot(data = tmp, aes(x = date, y = value)) + theme_bw() + 
       theme(axis.title = element_text(size = 16),
             axis.text= element_text(size = 14),
             axis.line = element_line(size = 1),
             legend.position = c(0.1, 0.85)) +
       xlab("date") + 
       ylab("indices (scaled to 1996-01-02)") + 
       scale_x_date(minor_breaks = "1 year") +
       scale_y_log10(breaks = c(1, 2, 4, 8)) +  
       geom_hline(h = 0, lty = 2, col = "orange") +
       geom_vline(xintercept = as.numeric(as.Date("1996-01-02")), lty = 2, col = "orange") +
       geom_line(aes(color = name)) + 
       scale_colour_manual(values = c("red2", "blue2", "green4"), name = "Index", 
                           labels = c("NASDAQ100", "S&P500", "Wilshire5000"))
```


<hr class="separator">
<a name="APPENDIX"></a>

# APPENDIX

<a name="APPENDIX-functions"></a>

## Functions

Source code of the functions defined in `scripts/my_functions.R`:

```{r read_my_functions, echo = FALSE, cache = FALSE}
read_chunk("./scripts/my_functions.R")
```

```{r print_chunk, eval = FALSE, cache = FALSE}
<<my_defs>>
```


<a name="SessionInfo"></a>

## R Session Info

```{r R_session_info}
sessionInfo()
```
---
