---
title       : Buy vs. Rent App
subtitle    : a simulation-based calculator
author      : Giovanni Fossati
job         : 
output  : 
  html_document:
    self_contained: false
---

```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
require(knitr)
options(width = 100, 
        scipen = 1)
opts_chunk$set(message = FALSE, 
               error = FALSE, 
               warning = FALSE, 
               collapse = TRUE, 
               tidy = FALSE,
               cache = FALSE, 
               cache.path = '.cache/', 
               comment = '#',
               fig.align = 'center', 
               dpi = 100, 
               fig.path = 'figures/')

library("dplyr")
library("tidyr")
library("magrittr")
library("ggplot2")
```

```{r load_packages, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
source("./scripts/my_functions_1.R")
source("./scripts/my_functions_2.R")
```

## What is it and why we want it

The _Buy vs. Rent_ question confuses countless people daily.   
We seek help online, being almost invariably mislead by over-simplistic tools.

The plethora of calculators found on the web in the majority of cases do not properly handle
several important aspects of the cost/benefit analysis of buying vs. renting. 

For instance:

* tax-related items, namely the tax benefits of the mortgage interest deduction, also considering 
  the fact that applying it means itemizing deductions, hence losing the _standard deduction_.
* the benefits of re-investing money potentially saved by renting instead of buying,
* the benefit of the return of investment of the capital not put into a down-payment.

<span style="color: #006600 !important;">
[This calculator](https://pedrosan.shinyapps.io/buy_vs_rent/) implements a more comprehensive 
and realistic picture, including these and other ingredients, and yields 
a __statistical assessment__ of the __buy vs. rent__ question __by means of simulations__.
</span>

--- 

## A Simulations-based Approach

Given a set of input parameters, 250 simulations are performed, with stochastic "predictions" 
of the time varying values of: 

* property appreciation, 
* return of investment of the "unused cash" when renting,
* inflation, 
* rent increase (over inflation).

For the first three, values are drawn from Normal distributions with means and standard deviations
given among the application inputs. 
The additional rent increase rate is drawn from an exponential distribution.

For each simulation a 'trade-off' value is computed, giving the difference between buying the
given property and renting (including the return of the investment of the cash not put into the
property) for each year, for the duration of the mortgage loan.
Positive values are in favor of buying, negative indicate that renting would be more beneficial financially.


--- 

## Application Results

The simulations results are summarized in three plots showing:

1. the trends of the 'tradeoff' amount, 
2. the fraction of simulations favoring buying over renting, over time,
3. the distribution of tradeoff amounts over time, highligthing the distributions at 1/2, 3/4 and at the end of the loan period.

The simulations performed by this simple Application are purely based on general probability density functions, 
namely Gaussian and Exponential.

However it is clear that for the _Annual Variations_ inputs it would be possible to apply a more
real-data-driven approach, by sampling their values from distributions derived from real data from
stock market or bonds investment returns, real estate property values, and inflation.

Investment returns could be drawn from a proxy for different investing styles: bond indexes for
conservative, S&P500 for middle-ground, NASDAQ composite for more aggressive approach.   
Real estate property values could be adjusted regionally, increasing the usefulness and relevance of the Calculator.

The data exist and accessible.  I just did not have time to bring them into _the fold_... 

An example is shown below.

#### Example of Summary Plot

<button class="toggle_code">show output</button>
```{r prepare_for_plot, echo = TRUE}
# Define parameters (names and values)

paramNames <- c("start_prop_value", "down_payment_pct", "mortgage_rate", "n_years", 
                "initial_fixed_costs",
                "prop_tax_rate_pct", "prop_insurance", "HOA_monthly_fee",
                "start_rent", "rent_insurance",
                "annual_appreciation", "annual_appreciation_sd",
                "annual_inv", "annual_inv_sd",
                "annual_inflation", "annual_inflation_sd",
                "annual_rent_extra_increase_mean",
                "fraction_extra_cash_invested_pct",
                "income_tax_rate_pct", "itemized_deductions", "std_deduction",
                "n_sim")

fake_input <- list(start_prop_value = 200000, 
                   down_payment_pct = 20.0, 
                   mortgage_rate = 4.5, 
                   n_years = 30, 
                   initial_fixed_costs = 6000, 
                   prop_tax_rate_pct = 2.05,
                   prop_insurance = 4000.0, 
                   HOA_monthly_fee = 400.0,
                   start_rent = 1500.0, 
                   rent_insurance = 260.0, 
                   annual_appreciation = 3.0, 
                   annual_appreciation_sd = 2.0, 
                   annual_inv = 5.0, 
                   annual_inv_sd = 7.0, 
                   annual_inflation = 1.5, 
                   annual_inflation_sd = 1.0, 
                   annual_rent_extra_increase_mean = 0.5, 
                   fraction_extra_cash_invested_pct = 50.0, 
                   income_tax_rate_pct = 25.0, 
                   itemized_deductions = 0.0, 
                   std_deduction = 12200, 
                   n_sim = 200)
    
sims.tradeoff <- do.call(simulate_tradeoff, getParams1(fake_input, paramNames))

n.sim <- getParams1(fake_input, "n_sim")[[1]]
n.years <- getParams1(fake_input, "n_years")[[1]]
```

<button class="toggle_plot_code">show plot code</button>
```{r plot_results, cache = FALSE, fig.width = 9.0, fig.height = 6.0, echo = TRUE}
plot_sims(n.sim, n.years, sims.tradeoff)
```
    
## Model Parameters

### Buy, Rent and Tax Parameters

#### Property and mortgage

* __Purchase price__ (\$)
* __Down payment__ (%)
* __Mortgage interest rate__ (%)
* __Duration__ (years)
* __Initial fixed costs__ (\$): additional cost incurred when buying a property, e.g. closing costs, or repairs.
       
#### Ownership costs: property taxes, insurance, fees

* __Property tax rate__ (%)
* __Insurance cost__ (\$): home-owner insurance premium (annual).
* __HOA fee__ (\$): home-owner association fees (monthly).
          
#### Rent

* __Rent__ (\$): monthly, ideally for a property comparable to that considered for purchase.
* __Renter insurance__ (\$)
* __Fraction of saved cash re-invested__ (%): 
    if the total costs of renting are lower than those of owning, a portion of the saved cash can be re-invested.
    This parameter regulates the fraction of this saved cash that is added to the investments.
    
#### Income Tax Related

* __Marginal income tax__ (%): tax rate to use to calculate the potential tax-savings of the deduction of mortgage interests.
* __Other itemized deductions__ (\$): the mortgage interest deduction can only be taken if one itemizes all deductions, thus
	        losing the standard deduction. Because of this, the actual benefit of the mortgage interest deduction is only 
                related to the portion that exceeds the standard deduction.
* __Standard deduction__: please note that it may be different if filing as married or separately.


### Annual Variations

#### Property appreciation

Assuming uncorrelated normally distributed values.

* __Appreciation__ (%): mean yearly increase of property values.
* __Appreciation std. dev.__ (%): "volatility" of the property value changes.
          
#### Cash investment return
          
Assuming uncorrelated normally distributed values.

* __Return__ (%):  mean yearly return of cash investments.
* __Return std. dev.__ (%): "volatility" of cash investment returns.

#### Inflation

Assuming uncorrelated normally distributed values.

* __Inflation__ (%): mean yearly inflation rate.
* __Inflation std. dev.__ (%): "volatility" of inflation.
          
#### Rent increase

* __Extra increase over inflation__ (%): extra rate of increase of rent, on top of inflation.   
      Values are drawn from an exponential distribution with this mean.


<hr class="thin_separator">
<a name="APPENDIX"></a>

# APPENDIX

<a name="APPENDIX-functions"></a>

## Functions

Source code of the user-defined functions in 
[scripts/my_functions.R](https://github.com/pedrosan/DataScienceExamples/tree/master/Rent_vs_Buy/Returns/scripts/my_functions.R) 


```{r read_my_functions, eval = TRUE, echo = FALSE, cache = FALSE}
read_chunk("./scripts/my_functions_1.R")
read_chunk("./scripts/my_functions_2.R")
```

<button class="toggle_code">show code</button>
```{r print_chunk_1, eval = FALSE, echo = TRUE, cache = FALSE}
<<my_functions_1>>
```

<button class="toggle_code">show code</button>
```{r print_chunk_2, eval = FALSE, echo = TRUE, cache = FALSE}
<<my_functions_2>>
```


<a name="SessionInfo"></a>

## R Session Info

```{r R_session_info}
sessionInfo()
```

---
