---
title: "Stock Market Returns Distributions"
subtitle : 
author   : Giovanni Fossati
job      : null
output   : 
  html_document:
    self_contained: false
    css: css/gf_new.css
    theme: cerulean
    highlight: tango
    keep_md: no
    mathjax: default
    toc: no
    includes:
      md_extensions: +fenced_code_attributes
---

```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
require("knitr")
options(width = 100, 
        scipen = 5)
opts_chunk$set(message = FALSE, 
               error = FALSE, 
               warning = FALSE, 
               collapse = TRUE, 
               tidy = FALSE,
               cache = FALSE, 
               cache.path = '.cache/', 
               comment = '#',
               fig.align = 'center', 
               dpi = 100, 
               dev = "png",
               fig.path = 'figures/')
# library("pkgmaker")
# knit_hooks$set(toggle = hook_toggle())
```

## PRELIMINARIES

Libraries needed for data processing and plotting:

```{r load_packages, cache = FALSE, echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE}
# library("lubridate")
# library("plyr")
library("ggplot2")
library("gridExtra")
```


## Introduction

Two interesting, blog posts:

* [The Generalized Lambda Distribution and GLDEX Package: Fitting Financial Return Data](http://www.r-bloggers.com/the-generalized-lambda-distribution-and-gldex-package-fitting-financial-return-data/)
* [The Generalized Lambda Distribution and GLDEX Package for Fitting Financial Return Data – Part 2](http://www.r-bloggers.com/the-generalized-lambda-distribution-and-gldex-package-for-fitting-financial-return-data-part-2/)

The GLDEX package is available on [CRAN](http://cran.r-project.org/web/packages/GLDEX/index.html)


<a name="THE_DATA"></a>

## Libraries and Other Setup

```{r load-libraries, echo = TRUE, cache = FALSE}
library("quantmod")
library("GLDEX")
library("MASS")
```

```{r load-more_libraries_for_plotting, echo = TRUE, cache = FALSE}
library("scales")
source("./my_functions.R")
```

```{r set_dates}
start.date <- "1986-01-01"
end.date <- "2014-12-31"

xxlin <- seq(-25.0, 25.0, by = 0.1)
xxlog <- seq(-2.0, 2.0, by = 0.001)
```

<hr class="thin_separator">

# Three Broad Market Indices : Daily Returns

## Wilshire 5000

Fetch the data with `quantmod`:

```{r wilshire5000-load, cache = TRUE}
# getSymbols("VTSMX", from = "1994-01-01")
getSymbols("VTSMX", from = start.date, to = end.date)
VTSMX.vec <- as.vector(VTSMX[, 4])
```

Prepare returns:

```{r wilshire5000-prepare_data, cache = TRUE}
WW <- prepare_data(data = VTSMX.vec, xlin = xxlin, xlog = xxlog)
WW$dates <- index(VTSMX)
WW$year <- substr(index(VTSMX), 1, 4)
```

```{r wilshire5000-plot_distributions_plain, echo = FALSE, cache = TRUE, eval = FALSE}
hist(WW$rr_data, breaks = seq(-55.0, 55.0, by = 0.10), xlim = c(-8.0, 8.0), freq = FALSE,
     xlab = "daily returns (%)",
     main = "Wilshire 5000")
lines(WW$rr_distr$x, WW$rr_distr$y_norm1, col = "green2", lwd = 2)
lines(WW$rr_distr$x, WW$rr_distr$y_norm2, col = "blue2", lwd = 2)
lines(WW$rr_distr$x, WW$rr_distr$y_gl, col = "red2", lwd = 2)
```

#### Distributions of _relative returns_ and of _log returns_:

```{r wilshire5000-plot_distributions_ggplot_3, echo = FALSE, cache = TRUE, fig.width = 10.0, fig.height = 5.0}
p1 <- plot_distr_rel_returns(data = WW$rr_data, fits = WW$rr_distr)
p2 <- plot_distr_log_returns(data = WW$rl_data, fits = WW$rl_distr)
grid.arrange(p1, p2, nrow=1)
```


<hr class="thin_separator">

### S&P 500 

Fetch the data with `quantmod`:

```{r sp500-load, cache = TRUE}
getSymbols("^GSPC", from = start.date, to = end.date)
GSPC.vec <- as.vector(GSPC[, 4])
```

```{r sp500-prepare_data, cache = TRUE}
SP <- prepare_data(data = GSPC.vec, xlin = xxlin, xlog = xxlog)
SP$dates <- index(GSPC)
SP$year <- substr(index(GSPC), 1, 4)
```

#### Distributions of _relative returns_ and of _log returns_:

```{r sp500-plot_distributions_ggplot_3, echo = FALSE, cache = TRUE, fig.width = 10.0, fig.height = 5.0}
p1 <- plot_distr_rel_returns(data = SP$rr_data, fits = SP$rr_distr)
p2 <- plot_distr_log_returns(data = SP$rl_data, fits = SP$rl_distr)
grid.arrange(p1, p2, nrow=1)
```


<hr class="thin_separator">

### NASDAQ 100

Fetch the data with `quantmod`:

```{r nasdaq100-load, cache = TRUE}
getSymbols("^NDX", from = start.date, to = end.date)
NDX.vec <- as.vector(NDX[, 4])
```

```{r nasdaq100-prepare_data, cache = TRUE}
NN <- prepare_data(data = NDX.vec, xlin = xxlin, xlog = xxlog)
NN$dates <- index(NDX)
NN$year <- substr(index(NDX), 1, 4)
```

#### Distributions of _relative returns_ and of _log returns_:

```{r nasdaq100-plot_distributions_ggplot_3, echo = FALSE, cache = TRUE, fig.width = 10.0, fig.height = 5.0}
p1 <- plot_distr_rel_returns(data = NN$rr_data, fits = NN$rr_distr)
p2 <- plot_distr_log_returns(data = NN$rl_data, fits = NN$rl_distr)
grid.arrange(p1, p2, nrow=1)
```


<hr class="thin_separator">

# Three Broad Market Indices : Daily Returns


## Annual from Daily Returns

Simulating annual returns by compounding daily returns drawn from the 
best fit _generalized gamma distributions_.

```{r more_code, eval = TRUE, cache = TRUE}
# typical number of market days in a year is ~252
ndays <- 252

# repeats
  N <- 1e5
# N <- 1e6

set.seed(1212)
test_rr.W5000 <- rep(0, N)
for(j in 1:N) {
    dummy <- rgl(ndays, WW$rr_fit_gl)
    test_rr.W5000[j] <- (cumprod(1 + dummy/100)[ndays] - 1.0)*100
}

test_rr.SP500 <- rep(0, N)
for(j in 1:N) {
    dummy <- rgl(ndays, SP$rr_fit_gl)
    test_rr.SP500[j] <- (cumprod(1 + dummy/100)[ndays] - 1.0)*100
}

test_rr.N100 <- rep(0, N)
for(j in 1:N) {
    dummy <- rgl(ndays, NN$rr_fit_gl)
    test_rr.N100[j] <- (cumprod(1 + dummy/100)[ndays] - 1.0)*100
}
```

```{r plot-annual_returns_distr_1, echo = FALSE, eval = TRUE, fig.width = 8.0, fig.height = 5.5}
xx2 <- seq(-200, 200, by = 1.0)
yy.W5000.n <- dnorm(xx2, mean = mean(test_rr.W5000), sd = sd(test_rr.W5000))
yy.SP500.n <- dnorm(xx2, mean = mean(test_rr.SP500), sd = sd(test_rr.SP500))
yy.N100.n  <- dnorm(xx2, mean = mean(test_rr.N100), sd = sd(test_rr.N100))

plot(density(test_rr.N100), type = "n",
     xlim = c(-100.0, 200.0), 
     ylim = c(0.0, 0.037), 
     xlab = "annual return [%]", 
     main = "Wilshire5000, S&P500, NASDAQ100")
grid()
lines(xx2, yy.N100.n, col = "pink2", lty = 2, lwd = 3)
lines(xx2, yy.SP500.n, col = "lightblue", lty = 2, lwd = 3)
lines(xx2, yy.W5000.n, col = "lightgreen", lty = 2, lwd = 3)
lines(density(test_rr.N100), col = "red2", lwd = 3)
lines(density(test_rr.SP500), col = "blue2", lwd = 3)
lines(density(test_rr.W5000), col = "forestgreen", lwd = 3)
```

## Annual straight from data

```{r aligned_annual_returns, echo = FALSE}
WWb <- VTSMX.vec[WW$year >= 1994]
SPb <- GSPC.vec[SP$year >= 1994]
NNb <- NDX.vec[NN$year >= 1994]
WWd <- index(VTSMX)[WW$year >= 1994]
SPd <- index(GSPC)[SP$year >= 1994]
NNd <- index(NDX)[NN$year >= 1994]

W5000_rla <- diff(log(WWb), lag = 252)
W5000_rra <- 100.0*(exp(W5000_rla)-1)

SP500_rla <- diff(log(SPb), lag = 252)
SP500_rra <- 100.0*(exp(SP500_rla)-1)

N100_rla <- diff(log(NNb), lag = 252)
N100_rra <- 100.0*(exp(N100_rla)-1)

rr.df <- data.frame(date = WWd[-(1:252)], WW = W5000_rra, SP = SP500_rra, NN = N100_rra)
rr.df$time <- as.POSIXct(rr.df$date)

rr.df$WWidx <- WWb[-(1:252)]
rr.df$SPidx <- SPb[-(1:252)]
rr.df$NNidx <- NNb[-(1:252)]

rr.df$WWidx_n <- WWb[-(1:252)]/rr.df$WWidx[rr.df$date == "1996-01-02"]
rr.df$SPidx_n <- SPb[-(1:252)]/rr.df$SPidx[rr.df$date == "1996-01-02"]
rr.df$NNidx_n <- NNb[-(1:252)]/rr.df$NNidx[rr.df$date == "1996-01-02"]
```

```{r annual_returns, echo = FALSE, eval = FALSE}
W5000_rla <- diff(log(VTSMX.vec), lag = 252)
W5000_rra <- 100.0*(exp(W5000_rla)-1)

SP500_rla <- diff(log(GSPC.vec), lag = 252)
SP500_rra <- 100.0*(exp(SP500_rla)-1)

N100_rla <- diff(log(NDX.vec), lag = 252)
N100_rra <- 100.0*(exp(N100_rla)-1)
```

```{r plot-annual_returns_distr_2, echo = FALSE, eval = TRUE, fig.width = 8.0, fig.height = 5.5}
plot(density(W5000_rra), type = "n", 
     xlim = c(-100.0, 200.0), 
     ylim = c(0.0, 0.037), 
     xlab = "annual return [%]", 
     main = "Wilshire5000, S&P500, NASDAQ100")
grid()
lines(density(N100_rra), col = "red", lty = 2, lwd = 3)
lines(density(SP500_rra), col = "blue", lty = 2, lwd = 3)
lines(density(W5000_rra), col = "green2", lty = 2, lwd = 3)
lines(density(test_rr.N100), col = "red2", lwd = 3)
lines(density(test_rr.SP500), col = "blue2", lwd = 3)
lines(density(test_rr.W5000), col = "green3", lwd = 3)
```

```{r plot-annual_returns_timeseries, echo = FALSE, eval = FALSE, fig.width = 8.0, fig.height = 5.5}
plot(1:length(N100_rra), N100_rra, type = "n", 
     ylim = c(-75, 125),
     xlab = "index",
     ylab = "Annual Returns",
     main = "Wilshire5000, S&P500, NASDAQ100")
grid()
lines(1:length(N100_rra), N100_rra, type = "l", col = "red2")
lines(1:length(SP500_rra), SP500_rra, type = "l", col = "blue2")
lines(1:length(W5000_rra), W5000_rra, type = "l", col = "green3")
```

```{r plot-annual_returns_timeseries_ggplot, echo = FALSE, eval = TRUE, fig.width = 10.0, fig.height = 5.5}
ggplot(data = rr.df, aes(x = time)) + theme_bw() + 
       theme(axis.title = element_text(size = 16),
             axis.text= element_text(size = 14),
             axis.line = element_line(size = 1)) +
       ylim(-75.0, 125.0) +
       xlab("date") + 
       ylab("annualized returns") + 
       # scale_x_continuous(minor_breaks = seq(1994, 2015, by = 1)) + 
       # scale_x_date(minor_breaks = seq(1994, 2015, by = 1)) + 
       scale_x_date(minor_breaks = "1 year") +
       geom_hline(h=0, lty = 2, col = "orange") +
       geom_line(aes(x = date, y = NN), col = "red2") + 
       geom_line(aes(x = date, y = SP), col = "blue2") + 
       geom_line(aes(x = date, y = WW), col = "green4")
```

```{r plot-indices_timeseries_ggplot, echo = FALSE, eval = FALSE, fig.width = 10.0, fig.height = 5.5}
ggplot(data = rr.df, aes(x = time)) + theme_bw() + 
       theme(axis.title = element_text(size = 16),
             axis.text= element_text(size = 14),
             axis.line = element_line(size = 1)) +
       # ylim(-75.0, 125.0) +
       xlab("date") + 
       ylab("indices") + 
       # scale_x_continuous(minor_breaks = seq(1994, 2015, by = 1)) + 
       # scale_x_date(minor_breaks = seq(1994, 2015, by = 1)) + 
       scale_x_date(minor_breaks = "1 year") +
       scale_y_log10() +  
       geom_line(aes(x = date, y = NNidx), col = "red2") + 
       geom_line(aes(x = date, y = SPidx), col = "blue2") + 
       geom_line(aes(x = date, y = WWidx), col = "green4")
```

```{r plot-indices_timeseries_ggplot_norm, echo = FALSE, eval = TRUE, fig.width = 10.0, fig.height = 5.5}
ggplot(data = rr.df, aes(x = time)) + theme_bw() + 
       theme(axis.title = element_text(size = 16),
             axis.text= element_text(size = 14),
             axis.line = element_line(size = 1)) +
       # ylim(-75.0, 125.0) +
       xlab("date") + 
       ylab("indices (scaled to 1996-01-02)") + 
       # scale_x_continuous(minor_breaks = seq(1994, 2015, by = 1)) + 
       # scale_x_date(minor_breaks = seq(1994, 2015, by = 1)) + 
       scale_x_date(minor_breaks = "1 year") +
       scale_y_log10(breaks = c(1, 2, 4, 8)) +  
       geom_line(aes(x = date, y = NNidx_n), col = "red2") + 
       geom_line(aes(x = date, y = SPidx_n), col = "blue2") + 
       geom_line(aes(x = date, y = WWidx_n), col = "green4")
```


<hr class="separator">
<a name="APPENDIX"></a>

# APPENDIX

<a name="SessionInfo"></a>

## R Session Info

```{r R_session_info}
sessionInfo()
```
---
